---
title: "spongeEffects.Rmd"
author: "Hoffmann M"
date: "2/10/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## (A) Loading dependencies for spongEffects

We start with loading the package and its dependencies

```{r, warning=FALSE, message=FALSE}
library(SPONGE)
library(doParallel)
library(foreach)

#if you want to use parallelization, you can register your backend here
num.of.cores <- 25 #many more on a compute cluster
cl <- makeCluster(num.of.cores) 
registerDoParallel(cl)

#will be deleted after testing
library(biomaRt)
library(dplyr)
source("/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/R/fn_spongeffects_utility.R")
load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/train_cancer_gene_expr.RDATA")
load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/train_cancer_mir_expr.RDATA")
load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/train_cancer_metadata.RDATA")

load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/test_cancer_gene_expr.RDATA")
load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/test_cancer_mir_expr.RDATA")
load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/test_cancer_metadata.RDATA")

load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/spongEffects_targets.RDATA")

load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/spongEffects_ceRNA_interactions_sign.RDATA")
load(file="/nfs/data/TCGA_TGCT/spongEffects_package/SPONGE/data/spongEffects_network_centralities.RDATA")
```

## (B) Formats of input data necessary for spongEffect

spongEffects comes with a very small example gene and miRNA expression dataset useful for illustrating functionality. We provide gene and miRNA expression datasets for train and test data. We also provide a small ceRNA network, gene-miRNA targets, centrality measures (of the ceRNA network), and a small metadata for this example (the ceRNA network and the centrality measures were created using the SPONGE vignette).

After loading the package, the example datasets can be accessed:

### Gene expression:
#### Train:
```{r, eval=FALSE}
head(train_cancer_gene_expr)
```

```{r, echo=FALSE, results='asis'}
knitr::kable(train_cancer_gene_expr[1:5,1:8])
```
#### Test:
```{r, eval=FALSE}
head(test_cancer_gene_expr)
```

```{r, echo=FALSE, results='asis'}
knitr::kable(test_cancer_gene_expr[1:5,1:8])
```

### miRNA expression:
#### Train
```{r, eval=FALSE}
head(train_cancer_mir_expr)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(train_cancer_mir_expr[1:5,1:8])
```
#### Test
```{r, eval=FALSE}
head(test_cancer_mir_expr)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(test_cancer_mir_expr[1:5,1:8])
```

### ceRNA network
```{r, eval=FALSE}
head(ceRNA_interactions)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(spongEffects_ceRNA_interactions_sign[1:5,1:8])
```

### gene-miRNA-targets
```{r, eval=FALSE}
head(spongEffects_targets)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(spongEffects_targets[1:5,1:8])
```

### centrality measures (of the ceRNA network)
```{r, eval=FALSE}
head(spongEffects_network_centralities)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(spongEffects_network_centralities[1:5,1:5])
```

### specification of meta data (e.g., subtypes of a cancer, column name of the subtypes is important) (format name = TCGA)
#### Train
```{r, eval=FALSE}
head(train_cancer_metadata)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(train_cancer_metadata[1:5,1:8])
```

#### Test
```{r, eval=FALSE}
head(test_cancer_metadata)
```

```{r, warning=FALSE, message=FALSE}
knitr::kable(test_cancer_metadata[1:5,1:8])
```

## (C) Filter ceRNA network

We now want to filter the network to only use significant ceRNA interactions

```{r message=FALSE, warning=FALSE}
filtered_network_centralities=filter_ceRNA_network(sponge_effects = spongEffects_ceRNA_interactions_sign, network_analysis = spongEffects_network_centralities, mscor.threshold = 0.01, padj.threshold = 0.1)
```

## (D) Discover modules

We now want to discover spongEffects modules, here you can choose, if you want lncRNAs, circRNAs, protein_coding or a combination of those classes to be the central of a module.
```{r, warning=FALSE, message=FALSE}

central_gene_modules<-get_central_modules(bioMart_gene_ensembl = "hsapiens_gene_ensembl", weighted_node_centrality = filtered_network_centralities$Node.Centrality,ceRNA_class = c("protein_coding"), cutoff = 100)

```

## (E) Define Modules 

Here, we search for the surrounding ceRNAs (all ceRNA classes are now considered).
```{r, warning=FALSE, message=FALSE}

Sponge.modules <- define_modules(network = filtered_network_centralities$Sponge.filtered, central.modules = central_gene_modules, remove.central = F, set.parallel = F)
# Module size distribution
Size.modules <- sapply(Sponge.modules, length)

```

## (F) Calculate Enrichment Scores (OE)
### Train:
```{r, warning=FALSE, message=FALSE}

#Either method="OE" (OverallEnrichment) or method="GSVA" (Gene Set Variation Analysis)
#We used OE in our Boniolo and Hoffmann et al., 2022

train.modules <- enrichment_modules(Expr.matrix = train_cancer_gene_expr, modules = Sponge.modules, bin.size = 10, min.size = 0, max.size = 2000, method = "OE")

```
### Test:
```{r, warning=FALSE, message=FALSE}

#Either method="OE" (OverallEnrichment) or method="GSVA" (Gene Set Variation Analysis)
#We used OE in our Boniolo and Hoffmann et al., 2022

test.modules <- enrichment_modules(Expr.matrix = test_cancer_gene_expr, modules = Sponge.modules, bin.size = 100, min.size = 10, max.size = 200, method = "OE")

```

## (F) 1. Train and test random forests
```{r, warning=FALSE, message=FALSE}
trained.model = train_and_test_model(Modules_training = train.modules, Modules_training.metadata = train_cancer_metadata, Modules_testing = test.modules, Modules_testing.metadata = test_cancer_metadata, Modules_testing.metadata.type ="TCGA", core = 1)

#TODO: Modules_testing.metadata.type set default to TCGA
```
## (J) 2. Train and test on Central Genes
```{r, warning=FALSE, message=FALSE}
central.genes.model = build_classifier_central_genes(expression.data.set1 = prep_list_tcga_spongEffects$TCGA.expr.tumor, expression.data.set2 = prep_list_METABRIC_spongEffects$METABRIC.expr, enrichment.modules.data.set1 = BRCA.Modules.OE, enrichment.modules.data.set2 = METABRIC.Modules.OE, meta.data.set1 = prep_list_tcga_spongEffects$TCGA.meta.tumor, meta.data.set2 = prep_list_METABRIC_spongEffects$METABRIC.meta, type.data.set1 = "TCGA", type.data.set2 = "METABRIC", Metric="Exact_match", n.folds=10)
```

## (J) 3. Randomly train and test
```{r, warning=FALSE, message=FALSE}
random.model = build_classifier_random(Sponge.modules= Sponge.modules,expression.data.set1 = prep_list_tcga_spongEffects$TCGA.expr.tumor, expression.data.set2 = prep_list_METABRIC_spongEffects$METABRIC.expr, meta.data.set1 = prep_list_tcga_spongEffects$TCGA.meta.tumor, meta.data.set2 = prep_list_METABRIC_spongEffects$METABRIC.meta, type.data.set1 = "TCGA", type.data.set2 = "METABRIC", Metric="Exact_match", n.folds=10)
```

## (K) Plot top k modules
```{r, warning=FALSE, message=FALSE}
lollipop_plot=plot_top_modules(trained.model=trained.model)
```


## (L) Density of module scores
```{r, warning=FALSE, message=FALSE}
density_plot_TCGA=plot_density_scores(trained.model=trained.model,modules = BRCA.Modules.OE,meta_data =prep_list_tcga_spongEffects$TCGA.meta.tumor,data_type="TCGA",subtypes = tcga_subtypes)
density_plot_METABRIC=plot_density_scores(trained.model=trained.model,modules = METABRIC.Modules.OE, meta_data = prep_list_METABRIC_spongEffects$METABRIC.meta,data_type = "METABRIC",subtypes = tcga_subtypes)
```

## (M) plot accuracy, sensitiviy, and specificity
```{r, warning=FALSE, message=FALSE}
metric_plot_list<-plot_accuracy_sensitivity_specificity(trained.model = trained.model,CentralGenes.model = central.genes.model,Random.model = random.model,training_dataset_name = "TCGA",testing_dataset_name = "METABRIC",subtypes = tcga_subtypes)

```

## (N) plot confusion matrices
```{r, warning=FALSE, message=FALSE}

subtypes.testing.factors <-prep_list_METABRIC_spongEffects$METABRIC.meta$CLAUDIN_SUBTYPE[complete.cases(t(METABRIC.Modules.OE))]

conf_matrix_trained<-plot_confusion_matrices(trained.model = trained.model, subtypes.testing.factors = subtypes.testing.factors)
conf_matrix_random<-plot_confusion_matrices(trained.model = random.model,subtypes.testing.factors = subtypes.testing.factors)
```

## (O) plot module heatmaps
## TODO: METABRIC NOT WORKING! HAVE TO FIND BUG
```{r, warning=FALSE, message=FALSE}

tcga.heatmap<-plot_heatmaps_training_test(trained.model = trained.model, Sponge.modules = Sponge.modules,enrichment.modules = BRCA.Modules.OE, meta.data = prep_list_tcga_spongEffects$TCGA.meta.tumor,type.data.set = "TCGA",subtypes = tcga_subtypes)
#metabric.heatmap<-plot_heatmaps_training_test(trained.model = trained.model, Sponge.modules = Sponge.modules,enrichment.modules = METABRIC.Modules.OE, meta.data = prep_list_METABRIC_spongEffects$METABRIC.meta, type.data.set = "METABRIC",subtypes = tcga_subtypes)

```

## (P) plot miRNA to modules heatmap
```{r, warning=FALSE, message=FALSE}
plot_involved_miRNAs_to_modules(Sponge.modules = Sponge.modules, trained.model = trained.model,dir_miRNAs_significance = "/nfs/data/SPONGE/results/breast invasive carcinoma_mirna_targets_all_mirs.RData")
```

## (Q) stop the cluster for parallelization
```{r, warning=FALSE, message=FALSE}
#stop your backend parallelisation if registered
stopCluster(cl) 
```